Seismic parameter estimation and the Canadian crust

### 2013-12-04 paper/draftBP.tex
intro revisions, still need to add eaton multigenic moho

%% -----------------------------------------------%%
\section{Introduction}

%% Seismic studies of the Canadian continental crust predominately analyze specific geological features or regions rather than taking a comprehensive and comparative inter-regional analysis. The primary reason for this is the poor resolution afforded by the seismic networks currently and historically deployed across Canada. The Canadian continental landmass is composed of at least fifteen large geological provinces as recognized by the Geological Survey of Canada (GSC). Each of these regions is itself complex and heterogeneous and often larger than most European nations. On top of this there is poor seismic coverage, roughly one seismic station per 25,000 $km^2$. Many of these stations are clustered near areas of geologic interest such as the Cascadia Subduction zone or population and research centres such as Southern Ontario or areas of significant resource interest like the diamondiferous Great Slave Lake area. This leaves vast areas of the Canadian landmass completely unsampled. Despite these limitations, a low resolution comprehensive and comparative study is still feasible for some of the geological regions comprising the Canadian continental crust. Direct comparison between the bulk average seismic properties of geological regions and between aggregated regions and global averages provide some new insight into the crustal composition of the Canadian landmass. The accumulated dataset also affords investigations into variations of bulk crustal parameters such as Poisson's Ratio and crustal thickness with age and tectonic environment as well as some basic statistical data on interesting crustal features.

%% This paper presents a comparative tour through the dataset accumulated from processing more than a decade worth of data from all available Canadian seismic stations. It begins with a discussion of the raw data itself followed by a review of the receiver function method and a more detailed explanation of the inversion algorithms used to produce the dataset. Three previous publications utilizing similar processing schemes provide unique subsets of data to compare with the values computed in this survey for quality assurance. ...

@@

@@

Efron, B., R. Tibshirani (1986), Bootstrap Methods for Standard Errors, Confidence Intervals, and Other Measures of Statistical Accuracy, Statistical Science, 1, 54-75.

@@

@@

@@

@@

>>>>

%% -----------------------------------------------%%
\section{Introduction}
\section{Introduction}

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior form part of the assemblage that comprises the cratonic core of Laurentia [Hammer et al., 2010]. Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield [Hoffman, 1988, Figure 1]. The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and sourthern Superior is composed of roughly eastwest trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6Ga [Thurston et al., 1991]. In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth [O'Neil et al., 2011]. To the east lies the younger, 1.2-0.98Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision [Eaton et al., 2006]. North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga [Thompson et al., 2010]. The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc [Kusky, 1989]. To the west of the Superior are the extensive Interior Platforms comprising the southern extents of the Trans-Hudson Orogen and Hearne Domain as well as the Alberta Orogen. The Interior platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism [Hammer et al., 2010]. Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurasic and Paleocene intense metamorphic, plutonic and uplifting events occured as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coastal mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock [Clowes et al, 1995]. Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale décollement beneath interleaved and duplexed crust [Hammer et al., 2010]. The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling [Clowes et al., 1999]. The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Oceans, and have enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of Cook et al [2009] who argue that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration. In another recent Lithoprobe study, Hammer and Clowes [2011] discuss broad structural characteristics of the Canadian crust such as their observation that Achean cratons and their boundaries contain structures very similar to those found in Proterozoic and Phanerozoic orogens. Another key observation is the lack of evidence for significant crustal disruption from magmatic instrusion related to subduction, post-orogenic extension or orogenic collapse. The authors also remark upon the striking uniformity in Moho depth that is imaged between 33km and 43km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. and Hammer and Clowes raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary. The Lith5.0 revised crustal model presented by Perry, Eaton and Forte [2002] was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model [Mooney et al, 1998] and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick (>42.5km) Phanerozoic crust in the south and west of the continental craton.

In this study, we examine the additional constraints that analyses of teleseismic receiver functions place upon Canadian bulk crustal properties when combined with available active source results. It is widely appreciated that receiver functions are sensitive primarily to crustal thickness and Poisson's ratio (or equivalently $V_P/V_S$) (Zhu and Kanomori, 20xx), and that the latter quantity provides some ability to discriminate among possible bulk compositions [Zandt and Ammon 1995]. The constraint on Poisson's ratio and the greater averaging performed by low frequency (0.04-3 Hz) scattered teleseismic waves versus high frequency (10-40 Hz) {\it P}-waves employed in active source studies render the two techniques somewhat complementary. Our data set comprises the majority of teleseismic data available for the Canadian landmass including both permanent and temporary deployments. Many of the temporary deployments are described in [Bostock, Eaton, Snyder]. Three previous studies employing similar methodologies but with more regional emphases deserve special mention: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen [Eaton et al., 2005], a structural and compositional characterization of the Superior Province [Darbyshire et al. 2006] and a crustal evolution analysis of the Hudson Bay region [Thompson et al., 2010]. Results from these studies will serve as benchmarks for our expanded effort. In the following sections we describe the data and methods employed to interrogate the Canadian crust. We proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three studies mentioned above for identical seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the resulting crustal property estimates.  First presented are the average crustal discontinuity and structural patterns across Canada. Second, we take a tour through those geological provinces for which sufficient data exists, focusing on regional variation and characterization. We conclude with a discussion on the implications for our understanding of the composition and evolution of the Canadian Crust.

@@

Clowes, R. M., F. Cook, Z. Hajnal, J. Hall, J. Lewry, S. Lucas, R. Wardle (1999), Canada's Lithoprobe Project (collaborative, multidisciplinary geoscience research leads to new understanding of continental evolution), Episodes, 22, 3-20.

Clowes, R. M., C. A. Zelt, J. R. Amor, R. M. Ellis (1995), Lithospheric structure in the southern Canadian Cordillera from a network of seismic refraction lines, Can. J. Earth Sci., 32, 1485-1513.

@@

Efron, B., R. Tibshirani (1986), Bootstrap methods for standard errors, confidence intervals, and other measures of statistical accuracy, Statistical Science, 1, 54-75.

@@

Hammer, P. T., R. M. Clowes (2011), The big picture: a lithospheric cross section of the North American continent, GSA Today, 21, 4-9.

@@

Hoffman, P. F. (1998), United plates of America, the birth of a craton: Early Proterozoic Assembly and Growth of Laurentia, Ann. Rev. Earth Planet, 16, 543-603.

@@

Kusky, T. M. (1989), Accretion of the Archean Slave province Geology, 17, 63-67.

@@

Thurston, P. C., H. R. Williams, R. H. Sutcliffe, G. M. Scott (Eds.) (1991), Geology of Ontario, Ontario Geological Survey Special, 4.
